\LARGE \bf Planning with Attitude

### 2020-10-14 paper/main.tex
Address Zac's comments

    More recently, these ideas have been extended to trajectory generation
    \cite{Zefran1998}, sample-based motion planning \cite{Zefran1999,Kuffner2004}, and
    optimal control. Approaches to optimal control with attitude states include
    analytical methods applied to satellites \cite{Spindler1998}, discrete mechanics
    \cite{Kobilarov2011,Kobilarov2014, Lee2008}, a combination of sampling-based planning
    and constrained trajectory optimization for satellite formations \cite{Garcia2005,
    Aoude2008}, projection operators \cite{Saccon2013}, or more general theory for
    optimization on manifolds \cites{watterson2018trajectory}. Nearly all of these
    methods rely heavily on principles from differential geometry and Lie group theory;
    However, despite these works, many recent papers in the robotics community continue
    to naively apply standard methods for motion planning and control with no regard for the
    group structure of rigid body motion
    %\cite{Alothman2016,deCrousaz2015, Williams2017,Geisert2016}
    .

    In this paper, we make a departure from previous approaches to geometric planning and 
    control that rely heavily on ideas and notation from differential geometry, 
    and instead use only basic mathematical tools from linear algebra and vector calculus that 
    should be familiar to most roboticists. In Sec. \ref{sec:Quaternion_Calculus} we introduce 
    an approach to quaternion differential calculus similar to \cite{Mandic2011,Xu2016}, but significantly simpler and more general, enabling straight-forward adaptation of 
    existing algorithms to systems with quaternion states. 
    For concreteness, we then apply our method to the canonical
    Wahba's problem~\cite{Wahba1965} in Sec. \ref{sec:Wahbas}, and demonstrate superior convergence to approaches that
    fail to properly account for the group structure. 
    In Sec. \ref{sec:trajopt} we extend these ideas to the problem of trajectory optimization,
    and detail modifications to ALTRO, a state-of-the-art constrained trajectory optimization
    solver, and demonstrate performance gains on several benchmark problems.
    %To our knowledge, there does not currently exist a solver that is capable of leveraging the unique Markovian structure of the fixed-horizon trajectory optimization problem while correctly accounting for the group structure of 3D rotations.
    In summary, our contributions include:

@@

        \centering
        \includegraphics[width=\columnwidth, height=5cm]{figures/wahba_convergence.tikz}
        \caption{Convergence comparison for Wahba's problem. By modfying Newton's method with the methods of section \ref{sec:Quaternion_Calculus},        quadratic convergence rates are achieved, while a na\"ive approach stalls after only a few itterations.}
        \label{fig:wahba_convergence}
    \end{figure}
    % \begin{figure}[H]
    %     \centering
    %     \includegraphics[width=\columnwidth, height=5cm]{figures/kevins_plots/convergence_plot.tikz}
    %     \caption{Convergence comparison for Wahba's problem. By performing Newton's method
    %     on the error quaternion and applying the result to the full quaternion we achieve
    %     quadratic convergence, whereas the more na\"ive approach doesn't converge to zero
    %     error.}
    %     \label{fig:wahba_convergence}
    % \end{figure}

@@

\section{Experiments} \label{sec:experiments}
    In this section we present several trajectory optimization problems for systems that
    undergo large changes in attitude: an airplane barrel roll, a quadrotor flip, and a
    satellite with flexible solar panels that must slew to a new orientation while
    avoiding a keep-out zone. All results were run on a desktop computer with an AMD
    Ryzen 2950x processor with 40 GB of RAM. All problems are run using ALTRO, first
    without any of the modifications presented in the current paper, analagous to the
    naive Newton's method in section \ref{sec:Wahbas} and labeled ``naive'', and then
    using the modifications listed in Sec. \ref{sec:trajopt} and the geodesic cost
    function described in Sec. \ref{sec:cost_functions}, labeled ``modified''. Timing
    results are summarized in Table \ref{tab:timing_results}. All experiments were solved
    to a constraint satisfaction tolerance of $10^{-5}$ and discretized with a 4th order
    Runge-Kutta integrator. Code for all experiments is available on
    \href{https://github.com/RoboticExplorationLab/PlanningWithAttitude}
    {GitHub\footnote{\url{https://github.com/RoboticExplorationLab/PlanningWithAttitude}}}.

@@

        A 180 degree barrel roll trajectory for a fixed-wing airplane was optimized. The airplane's dynamics model consists of the a simple rigid body as defined in Section
        \ref{sec:rigidbody_dynamics} with forces and torques due to lift and drag fit
        from wind tunnel data \cite{manchester2016udp}. The airplane was tasked to do a
        barrel roll by setting a high terminal cost for deviations from an upside-down attitude (see Fig.
        \ref{fig:barrellroll}). The solver was initialized with level flight trim
        conditions. The convergence of the different versions of ALTRO is compared in
        Fig. \ref{fig:c_max_convergence}. As expected, the modified version achieves
        better convergence and faster solve times compared to the na\"ive version since
        the expansions being provided to the algorithm more accurately capture the
        relationship between the attitude state and the goal and constraints.
\todo{Brian: you mention constraints in the text and have it on the y-axis in Fig. 4, but you don't state what these constraints are in the text.}

\todo{Brian: comment on the performance numbers in the text.}

		% \todo{If we're not going to plot it and actually do an experimental comparison, I vote for leaving this stuff out.}
        % We also compared the ``error quadratic'' cost function described in Section
        % \ref{sec:error_quadratic} to the geodesic cost function in Section
        % \ref{sec:geodesic}. The geodesic cost was more efficient, both in term of
        % iterations and computation time, taking only 36 iterations and 1.8 ms/iteration
        % vs 76 iterations and 7.6 ms/iteration for the error quadratic. Not only is the
        % geodesic cost function much cheaper to compute, it also converges much faster
        % than the error quadratic cost function. This could be due to the strong nonlinearities
        % introduced when computing the quaternion error, making it more difficult to
        % optimize.

@@

    \subsection{Quadrotor Flip}
        A 360 degree flip trajectory for a quadrotor was optimized with dynamics adapted
        from \cite{mellinger2012trajectory}. Four
        intermediary ``keyframes'' were used to encourage the quadrotor to reacj angles
        of \ang{90}, \ang{180}, \ang{270}, and \ang{360} around an approximately circular
        arc. The quadrotor was constrained to stay above the floor and move to a goal
        state 2 meters away in the $+y$ direction. The solver was initialized with a
        dynamically infeasible trajectory that linearly interpolates between the initial
        and final states, rotating the quadrotor around the x-axis a full \ang{360}.

\todo{Brian: little more detail on cost weights. Any control cost? If we used the same cost function form for all of these (i.e. geodesic on q + quadratic on everything else), let's write it at the beginning of the experiment section.}

	    Figure \ref{fig:quad_flip} shows snapshots of the trajectory as generated using
	    ALTRO. The original version of ALTRO, even after significant tuning efforts,
	    was not able to converge to the desired solution.
	    %It instead got halfway and then ``unwound'' 360 degrees and then continued rotating to the final orientation.
	    This behavior is common and expected when attempting optimization that does not
	    properly account for the group structure of rotations. It is also worth noting
	    that this problem could not be solved using any three-parameter attitude
	    representation, since it passes through the singularities at $90\degree,
	    180\degree$, and $360\degree$ associated with Euler angles, Rodrigues parameters, and
	    Modified Rodrigues Parameters, respectively.

	    \todo{Brian: you say the original version didn't work at all, but there are timing results for it in the table???}

        \begin{figure}[t]
            \centering
            \includegraphics[width=\columnwidth]{figures/quadflip.png}
            \caption{Snapshots of the quadrotor flip trajectory. The
                gree-colored quadrotors represent the state near t=0 s and the
                red-colored quadrotors represent the state near t=5.0 s
            }
            \label{fig:quad_flip}
        \end{figure}

@@

        We enforce the camera keep-out zone can with the following constraint,
        \begin{equation}
            \left(\inframe{W}{r_{sun}}\right)^T 
            \left(\toframe{W}{B}{A(q)} \, \inframe{B}{r_{cam}} \right) \leq \cos(40 ^\circ) ,
        \end{equation}
        where $^Br_{cam}$ is the camera line-of-sight unit vector in the body frame and $^Wr_{sun}$ is the unit vector pointing to the sun in the world frame. The attitudes that satisfy this constraint comprise a non-convex set, with the
        constraint itself being nonlinear in $q$.
        \begin{figure}[H]
            \centering
            \includegraphics{figures/kevins_plots/attitude_slew_plot.tikz}
            \caption{Visualization of the flexible spacecraft slew with a keep-out zone. Attitude is parameterized with a Rodrigues parameter to visualize the trajectory in three dimensions. The constraint surface represents attitudes where the camera line-of-sight is within 40$^\circ$ of the sun. The unconstrained solution violates this constraint, while the constrained solution is able to avoid the keep out zone.}
            \label{fig:keepout}
        \end{figure}
        % \todo{Talk a little about the results in the plot here.}

>>>>

    More recently, these ideas have been extended to trajectory generation
    \cite{Zefran1998}, sample-based motion planning \cite{Zefran1999,Kuffner2004}, and
    optimal control. Approaches to optimal control with attitude states include
    analytical methods applied to satellites \cite{Spindler1998}, discrete mechanics
    \cite{Kobilarov2011,Kobilarov2014, Lee2008}, a combination of sampling-based planning
    and constrained trajectory optimization for satellite formations \cite{Garcia2005,
    Aoude2008}, projection operators \cite{Saccon2013}, or more general theory for
    optimization on manifolds \cites{watterson2018trajectory}. Nearly all of these
    methods rely heavily on principles from differential geometry and Lie group theory;
    however, despite these works, many recent papers in the robotics community continue
    to naively apply standard methods for motion planning and control with no regard for the
    group structure of rigid body motion.
    %\cite{Alothman2016,deCrousaz2015, Williams2017,Geisert2016}

    In this paper, we make a departure from previous approaches to geometric planning and
    control that rely heavily on ideas and notation from differential geometry, and
    instead use only basic mathematical tools from linear algebra and vector calculus
    that should be familiar to most roboticists. In Sec. \ref{sec:Quaternion_Calculus} we
    introduce an approach to quaternion differential calculus similar to
    \cite{Mandic2011,Xu2016}, but significantly simpler and more general, enabling
    straight-forward adaptation of existing algorithms to systems with quaternion states.
    For concreteness, we then apply our method to the canonical Wahba's
    problem~\cite{Wahba1965} in Sec. \ref{sec:Wahbas}, and demonstrate superior
    convergence to approaches that fail to properly account for the group structure. In
    Sec. \ref{sec:trajopt} we extend these ideas to the problem of trajectory
    optimization, and detail modifications to ALTRO, a state-of-the-art constrained
    trajectory optimization solver, and demonstrate performance gains on several
    benchmark problems. With the modifications presented in this paper, ALTRO explicitly 
    leverages both the structure of the trajectory optimization problem as well as the 
    group structure of 3D rotations, making it uniquely well-suited to solving challenging
    problems with near real-time performance.

    %To our knowledge, there does not currently exist a solver that is capable of leveraging the unique Markovian structure of the fixed-horizon trajectory optimization problem while correctly accounting for the group structure of 3D rotations. 
    In summary, our contributions include:

@@

        \centering
        \includegraphics[width=\columnwidth, height=5cm]{figures/wahba_convergence.tikz}
        \caption{Convergence comparison for Wahba's problem. By modfying Newton's method
        with the methods of section \ref{sec:Quaternion_Calculus}, quadratic convergence
        rates are achieved, while a na\"ive approach stalls after only a few
        iterations.}
        \label{fig:wahba_convergence}
    \end{figure}

@@

\section{Experiments} \label{sec:experiments}
    In this section we present several trajectory optimization problems for systems that
    undergo large changes in attitude: an airplane barrel roll, a quadrotor flip, and a
    satellite with flexible solar panels that must slew to a new orientation while
    avoiding a keep-out zone.    
    All problems are run using ALTRO, first without any of the modifications presented in
    the current paper, analagous to the naive Newton's method in section \ref{sec:Wahbas}
    and labeled ``naive'', and then using the modifications listed in Sec.
    \ref{sec:trajopt} and the geodesic cost function described in Sec.
    \ref{sec:cost_functions}, labeled ``modified''. All cost functions were of the
    following form:
    \begin{align}
        \ell_\text{naive}(x, \bar{x}, Q) &= \half (x - \bar{x})^T Q (x - \bar{x}) \\
        \ell_\text{modified}(x, \bar{x}, Q) &= \half (x-\bar{x})^T \bar{Q} (x-\bar{x}) + w (1
            \pm \bar{q}^T q)
    \end{align}
    where $\bar{x}$ is the reference state and $\bar{Q} = \text{diag}(Q_r, \vec{0}_4,
    Q_v, Q_\omega)$, with $Q_r,Q_v,Q_\omega$ being the weights of $Q$ for position, and
    linear and angular velocity, respectively.

    Timing results are summarized in Table \ref{tab:timing_results}. All experiments were
    solved to a constraint satisfaction tolerance of $10^{-5}$ and discretized with a 3rd
    order Runge-Kutta integrator. All results were run on a desktop computer with an AMD
    Ryzen 2950x processor with 40 GB of RAM. Code for all experiments is available on
    \href{https://github.com/RoboticExplorationLab/PlanningWithAttitude}
    {GitHub\footnote{\url{https://github.com/RoboticExplorationLab/PlanningWithAttitude}}}.

@@

        A 180 degree barrel roll trajectory for a fixed-wing airplane was optimized. The
        airplane's dynamics model consists of the a simple rigid body as defined in
        Section \ref{sec:rigidbody_dynamics} with forces and torques due to lift and drag
        fit from wind tunnel data \cite{manchester2016udp}. The airplane was tasked to do
        a barrel roll by constraining the terminal state to upside-down (see Fig.
        \ref{fig:barrellroll}). To avoid issues with the double-cover and integration error,
        the following was used to enforce a terminal orientation to $\bar{q}$:
        \begin{equation}
            \frac{q_v}{\norm{q}} - \bar{q}_v
                \sign\left(\bar{q}^T \frac{q }{ \norm{q}}\right) = 0
        \end{equation}

        The solver was initialized with level flight trim
        conditions. The convergence of the different versions of ALTRO is compared in
        Fig. \ref{fig:c_max_convergence}. As expected, the modified version achieves
        better convergence and faster solve times compared to the na\"ive version since
        the expansions being provided to the algorithm more accurately capture the
        relationship between the attitude state and the goal and constraints. For this 
        relatively simple problem, we gained a modest 7\% improvement in runtime, despite
        the additional matrix multiplications when calculating the cost and constraint
        expansions.

@@

    \subsection{Quadrotor Flip}
        A 360 degree flip trajectory for a quadrotor was optimized with dynamics adapted
        from \cite{mellinger2012trajectory}. To encourage the flip, we specified a 
        ``waypoint'' cost function of the following form:
        \begin{multline}
                  \sum_{k \in \mathcal{N}} \ell(x_k, \hat{x}, \hat{Q}) 
                + \sum_{k \in \mathcal{W}} \ell(x_k, \bar{x}_k, Q_w) 
                + \half \sum_{k=1} u_k^T R u_k
        \end{multline}
        with $\mathcal{W} = \{45,51,55,75,101\}$, $\mathcal{N} = \{1:101\} \setminus
        \mathcal{W}$. The four intermediary ``waypoints'' were used to encourage the
        quadrotor to reach angles of \ang{90}, \ang{180}, \ang{270}, and \ang{360} around
        an approximately circular arc. The last waypoint was used to encourage the
        quadrotor to move towards the final goal. The quadrotor was constrained to stay
        above the floor and move to a goal state 2 meters away in the $+y$ direction. The
        solver was provided a dynamically infeasible initial trajectory that linearly
        interpolates between the initial and final states, rotating the quadrotor around
        the x-axis a full \ang{360}.

	    Figure \ref{fig:quad_flip} shows snapshots of the trajectory as generated using
	    ALTRO. The original version of ALTRO, even after significant tuning efforts, was
        not able to converge to the desired solution. 
	    This behavior is common and expected when attempting optimization that does not
	    properly account for the group structure of rotations. It is also worth noting
	    that this problem could not be solved using any three-parameter attitude
	    representation, since it passes through the singularities at $90\degree,
	    180\degree$, and $360\degree$ associated with Euler angles, Rodrigues parameters,
	    and Modified Rodrigues Parameters, respectively.

        \begin{figure}[ht]
            \centering
            \includegraphics[width=\columnwidth]{figures/quadflip.png}
            \caption{Snapshots of the quadrotor flip trajectory. The
                gree-colored quadrotors represent the state near t=0 s and the
                red-colored quadrotors represent the state near t=5.0 s
            }
            \label{fig:quad_flip}
        \end{figure}

@@

        We enforce the camera keep-out zone can with the following constraint,
        \begin{equation}
            \left(\inframe{W}{r_{sun}}\right)^T 
            \left(\toframe{W}{B}{A(q)} \, \inframe{B}{r_{cam}} \right) \leq \cos(40 ^\circ) ,
        \end{equation}
        where $^Br_{cam}$ is the camera line-of-sight unit vector in the body frame and $^Wr_{sun}$ is the unit vector pointing to the sun in the world frame. The attitudes that satisfy this constraint comprise a non-convex set, with the
        constraint itself being nonlinear in $q$.
        \begin{figure}[ht]
            \centering
            \includegraphics{figures/kevins_plots/attitude_slew_plot.tikz}
            \caption{Visualization of the flexible spacecraft slew with a keep-out zone. Attitude is parameterized with a Rodrigues parameter to visualize the trajectory in three dimensions. The constraint surface represents attitudes where the camera line-of-sight is within 40$^\circ$ of the sun. The unconstrained solution violates this constraint, while the constrained solution is able to avoid the keep out zone.}
            \label{fig:keepout}
        \end{figure}
        % \todo{Talk a little about the results in the plot here.}
